### 2022-01-25 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_metaweb_transfer@452e53e0b3d1fcd3d390982f752d836b2efed7ec 🚀
Deploying to gh-pages from @ PoisotLab/ms_metaweb_transfer@abd848de24d19518dcde87822d941a473bad53d5 🚀

\begin{enumerate}
    \item Despite their importance in many ecological processes,
collecting data and information on ecological interactions is an
exceedingly challenging task. For this reason, large parts of the world
have a data deficit when it comes to species interactions, and how the
resulting networks are structured. As data collection alone is unlikely
to be sufficient, community ecologists must adopt predictive methods.%
    \item We present a methodological framework that uses graph
embedding and transfer learning to assemble a predicted list of trophic
interactions of a species pool for which their interactions are unknown.
Specifically, we `learn' the information from a known interaction
network by inferring the latent traits of species and infer the latent
traits of a species pool for which we have no \emph{a priori}
interaction data based on their phylogenetic relatedness to species from
the known network. The latent traits can then be used to predict
interactions and construct an interaction network.%
    \item Here we assembled a metaweb for Canadian mammals derived from
interactions in the European food web, despite only 4\% of common
species being shared between the two locations. The results of the
predictive model are compared against databases of recorded pairwise
interactions, showing that we correctly recover 91\% of known
interactions.%
    \item The framework itself is robust even when the known network is
incomplete or contains spurious interactions making it an ideal
candidate as a tool for filling gaps when it comes to species
interactions. We provide guidance on how this framework can be adapted
by substituting some approaches or predictors in order to make it more
generally applicable.%
\end{enumerate}

@@

The latent variables are created by performing a truncated Singular
Value Decomposition (t-SVD; Halko et al., 2011) on the adjacency matrix.
SVD is an appropriate embedding of ecological networks, which has
recently been shown to both capture their complex, emerging properties
(Strydom, Dalla Riva, et al., 2021) and to allow highly accurate
prediction of the interactions within a single network (Poisot, Ouellet,
et al., 2021). Under SVD, an adjacency matrix \(\mathbf{A}\) (where
\(\mathbf{A}_{m,n}\in\mathbb{B}\) where 1 indicates predation and 0 an
absence thereof) is decomposed into three components resulting in
\(\mathbf{A} = \mathbf{U}\mathbf{\Sigma}\mathbf{V'}.\) Here,
\(\mathbf{\Sigma}\) is a \(m \times n\) diagonal matrix and contains
only singular (\(\sigma\)) values along its diagonal, \(\mathbf{U}\) is
a \(m \times m\) unitary matrix, and \(\mathbf{V}'\) a \(n \times n\)
unitary matrix. Truncating the SVD removes additional noise in the
dataset by omitting non-zero and/or smaller \(\sigma\) values from
\(\mathbf{\Sigma}\) using the rank of the matrix. Under a t-SVD
\(\mathbf{A}_{m,n}\) is decomposed so that \(\mathbf{\Sigma}\) is a
square \(r \times r\) diagonal matrix (with \(1 \le r \le r_{full}\)
where \(r_{full}\) is the full rank of \(\mathbf{A}\) and \(r\) the rank
at which we truncate the matrix) containing only non-zero \(\sigma\)
values. Additionally, \(\mathbf{U}\) is now a \(m \times r\) semi
unitary matrix and \(\mathbf{V}'\) a \(n \times r\) semi-unitary matrix.

The specific rank at which the SVD ought to be truncated is a difficult
question. The purpose of SVD is to remove the noise (expressed at high
dimensions) and to focus on the signal, (expressed at low dimensions).
In datasets with a clear signal/noise demarcation, a scree plot of
\(\mathbf{\Sigma}\) can show a sharp drop at the rank where noise starts
(Zhu \& Ghodsi, 2006). Because the European metaweb is almost entirely
known, the amount of noise (uncertainty) is low; this is reflected in
fig.~\ref{fig:scree} (left), where the scree plot shows no important
drop, and in fig.~\ref{fig:scree} (right) where the proportion of
variance explained increases smoothly at higher dimensions. For this
reason, we default back to a threshold that explains 60\% of the
variance in the underlying data, corresponding to 12 dimensions -
\emph{i.e.} a tradeoff between accuracy and a reduced number of
features.

An RDPG estimates the probability of observing interactions between
nodes (species) as a function of the nodes' latent variables, and is a
way to turn an SVD (which decompose one matrix into three) into two
matrices that can be multiplied to provide an approximation of the
network. The latent variables used for the RDPG, called the left and
right subspaces, are defined as
\(\mathscr{L} = \mathbf{U}\sqrt{\mathbf{\Sigma}}\), and
\(\mathscr{R} = \sqrt{\mathbf{\Sigma}}\mathbf{V}'\) -- using the full
rank of \(\mathbf{A}\), \(\mathscr{L}\mathscr{R} = \mathbf{A}\), and
using any smaller rank results in
\(\mathscr{L}\mathscr{R} \approx \mathbf{A}\). Using a rank of 1 for the
t-SVD provides a first-order approximation of the network. One advantage
of using an RDPG rather than an SVD is that the number of components to
estimate decreases; notably, one does not have to estimate the singular
values of the SVD. Furthermore, the two subspaces can be directly
multiplied to yield a network.

@@

It must be noted that despite bringing in a large amount of information
from the European species pool and interactions, the Canadian metaweb
has distinct structural properties. Following an approach similar to
Vermaat et al. (2009), we show in Supp. Mat. 3 that not only can we
observe differences in a multivariate space between the European and
Canadian metaweb, we can also observe differences in the same space
between random subgraphs from these networks. These results line up with
the studies spatializing metawebs that have been discussed in the
introduction: changes in the species pool are driving local structural
changes in the networks.

@@

\begin{figure}
\hypertarget{fig:inflation}{%
\centering
\includegraphics{figures/figure-validation.png}
\caption{Left, comparison of the probabilities of interactions assigned
by the model to all interactions (grey curve), the subset of
interactions found in GLOBI (red), and in the Strong \& Leroux (2014)
Newfoundland dataset (blue). The model recovers more interactions with a
low probability compared to data mining, which can suggest that
collected datasets are biased towards more common or easy to identify
interactions. Right, distribution of the in-degree and out-degree of the
mammals from Canada in the reconstructed metaweb. This figure describes
a flat, relatively short food web, in which there are few predators but
a large number of preys.}\label{fig:inflation}
}
\end{figure}

@@

Second, we looked for all species in the Canadian pool known to the
Global Biotic Interactions (GLoBI) database (Poelen et al., 2014), and
extracted their known interactions. Because GLoBI aggregates observed
interactions, it is not a \emph{networks} data source, and therefore the
only information we can reliably extract from it is that a species pair
\emph{was reported to interact at least once}. This last statement
should yet be taken with caution, as some sources in GLoBI (\emph{e.g.}
Thessen \& Parr, 2014) are produced through text analysis, and therefore
may not document direct evidence of the interaction. Nevertheless,
should the predictive model work, we would expect that a majority of
interactions known to GLoBI would also be predicted. We retrieved 366
interactions between mammals from the Canadian species pool from GLoBI,
33 of which were not predicted by the model; this results in a success
rate of 91\%. After performing this check, we set the probability of all
interactions known to GLoBI to 1.

Finally, we downloaded the data from Strong \& Leroux (2014), who mined
various literature sources to identify trophic interactions in
Newfoundland. This dataset documented 25 interactions between mammals,
only two of which were not part of our (Canada-level) predictions,
resulting in a success rate of 92\%. These two interactions were added
to our predicted metaweb with a probability of 1. A table listing all
interactions in the predicted Canadian metaweb can be found in the
supplementary material.

@@

Because the confidence intervals on the inferred trait space are
probably over-estimates, we decided to apply a thresholding step to the
interactions after data inflation (see fig.~\ref{fig:thresholds} showing
the effect of varying the cutoff on \(P(i \rightarrow j)\)). Cirtwill \&
Hambäck (2021) proposed a number of strategies to threshold
probabilistic networks. Their methods assume the underlying data to be
tag-based sequencing, which represents interactions as co-occurrences of
predator and prey within the same tags; this is conceptually identical
to our Bernoulli-trial based reconstruction of a probabilistic network.
We performed a full analysis of the effect of various cutoffs, and as
they either resulted in removing too few interactions, or removing
enough interactions that species started to be disconnected from the
network, we set this threshold for a probability equivalent to 0 to the
largest possible value that still allowed all species to have at least
one interaction with a non-zero probability. The need for this slight
deviation from the Cirtwill \& Hambäck (2021) method highlights the need
for additional development on network thresholding.

@@

Using a transfer learning framework we were able to construct a
probabilistic metaweb and (as per Dunne, 2006) is a list of potential
interactions, meaning that they will not necessarily exist wherever the
two species co-occur. The t-SVD embedding is able to learn relevant
ecological features for the network. fig.~\ref{fig:degree} shows that
the first rank correlates linearly with generality and vulnerability
(Schoener, 1989), \emph{i.e.} the number of preys and predators for each
species. Importantly, this implies that a rank 1 approximation
represents the configuration model for the metaweb, \emph{i.e.} a set of
random networks generated from a given degree sequence (Park \& Newman,
2004). Accounting for the probabilistic nature of the degrees, the rank
1 approximation also represents the \emph{soft} configuration model (van
der Hoorn et al., 2018). Both models are maximum entropy graph models
(Garlaschelli et al., 2018), with sharp (all network realizations
satisfy the specified degree sequence) and soft (network realizations
satisfy the degree sequence on average) local constraints, respectively.
The (soft) configuration model is an unbiased random graph model widely
used by ecologists in the context of null hypothesis significance
testing of network structure (\emph{e.g.} Bascompte et al., 2003) and
can provide informative priors for Bayesian inference of network
structure (\emph{e.g.} J.-G. Young et al., 2021). It is noteworthy that
for this metaweb, the relevant information was extracted at the first
rank. Because the first rank corresponds to the leading singular value
of the system, the results of fig.~\ref{fig:degree} have a
straightforward interpretation: degree-based processes are the most
important in structuring the mammalian food web.

@@

\textbf{Acknowledgements:} We acknowledge that this study was conducted
on land within the traditional unceded territory of the Saint Lawrence
Iroquoian, Anishinabewaki, Mohawk, Huron-Wendat, and Omàmiwininiwak
nations. TP, TS, DC, and LP received funding from the Canadian Institue
for Ecology \& Evolution. FB is funded by the Institute for Data
Valorization (IVADO). TS, SB, and TP are funded by a donation from the
Courtois Foundation. CB was awarded a Mitacs Elevate Fellowship no.
IT12391, in partnership with fRI Research, and also acknowledges funding
from Alberta Innovates and the Forest Resources Improvement Association
of Alberta. M-JF acknowledges funding from NSERC Discovery Grant and
NSERC CRC. RR is funded by New Zealand's Biological Heritage Ngā Koiora
Tuku Iho National Science Challenge, administered by New Zealand
Ministry of Business, Innovation, and Employment. BM is funded by the
NSERC Alexander Graham Bell Canada Graduate Scholarship and the FRQNT
master's scholarship. LP acknowledges funding from NSERC Discovery Grant
(NSERC RGPIN-2019-05771). TP acknowledges financial support from NSERC
through the Discovery Grants and Discovery Accelerator Supplement
programs.

>>>>

\begin{enumerate}
    \item Despite their importance in many ecological processes,
collecting data and information on ecological interactions is an
exceedingly challenging task. For this reason, large parts of the world
have a data deficit when it comes to species interactions, and how the
resulting networks are structured. As data collection alone is unlikely
to be sufficient, community ecologists must adopt predictive methods.%
    \item We present a methodological framework that uses graph
embedding and transfer learning to assemble a predicted list of trophic
interactions of a species pool for which their interactions are unknown.
Specifically, we `learn' the information (latent traits) of species from
a known interaction network and infer the latent traits of another
species pool for which we have no \emph{a priori} interaction data based
on their phylogenetic relatedness to species from the known network. The
latent traits can then be used to predict interactions and construct an
interaction network.%
    \item Here we assembled a metaweb for Canadian mammals derived from
interactions in the European food web, despite only 4\% of common
species being shared between the two locations. The results of the
predictive model are compared against databases of recorded pairwise
interactions, showing that we correctly recover 91\% of known
interactions.%
    \item The framework itself is robust even when the known network is
incomplete or contains spurious interactions making it an ideal
candidate as a tool for filling gaps when it comes to species
interactions. We provide guidance on how this framework can be adapted
by substituting some approaches or predictors in order to make it more
generally applicable.%
\end{enumerate}

@@

The latent variables are created by performing a truncated Singular
Value Decomposition (t-SVD; Halko et al., 2011) on the adjacency matrix.
SVD is an appropriate embedding of ecological networks, which has
recently been shown to both capture their complex, emerging properties
(Strydom, Dalla Riva, et al., 2021) and to allow highly accurate
prediction of the interactions within a single network (Poisot, Ouellet,
et al., 2021). Under SVD, an adjacency matrix \(\mathbf{A}\) (where
\(\mathbf{A}_{m,n}\in\mathbb{B}\) where 1 indicates predation and 0 an
absence thereof) is decomposed into three components resulting in
\(\mathbf{A} = \mathbf{U}\mathbf{\Sigma}\mathbf{V'}.\) Here,
\(\mathbf{\Sigma}\) is a \(m \times n\) diagonal matrix and contains
only singular (\(\sigma\)) values along its diagonal, \(\mathbf{U}\) is
a \(m \times m\) unitary matrix, and \(\mathbf{V}'\) a \(n \times n\)
unitary matrix. Truncating the SVD removes additional noise in the
dataset by omitting non-zero and/or smaller \(\sigma\) values from
\(\mathbf{\Sigma}\) using the rank of the matrix. Under a t-SVD
\(\mathbf{A}_{m,n}\) is decomposed so that \(\mathbf{\Sigma}\) is a
square \(r \times r\) diagonal matrix (with \(1 \le r \le r_{full}\)
where \(r_{full}\) is the full rank of \(\mathbf{A}\) and \(r\) the rank
at which we truncate the matrix) containing only non-zero \(\sigma\)
values. Additionally, \(\mathbf{U}\) is now an \(m \times r\) semi
unitary matrix and \(\mathbf{V}'\) an \(r \times n\) semi-unitary
matrix.

The specific rank at which the SVD ought to be truncated is a difficult
question. The purpose of SVD is to remove the noise (expressed at high
dimensions) and to focus on the signal (expressed at low dimensions). In
datasets with a clear signal/noise demarcation, a scree plot of
\(\mathbf{\Sigma}\) can show a sharp drop at the rank where noise starts
(Zhu \& Ghodsi, 2006). Because the European metaweb is almost entirely
known, the amount of noise (uncertainty) is low; this is reflected in
fig.~\ref{fig:scree} (left), where the scree plot shows no important
drop, and in fig.~\ref{fig:scree} (right) where the proportion of
variance explained increases smoothly at higher dimensions. For this
reason, we default back to a threshold that explains 60\% of the
variance in the underlying data, corresponding to 12 dimensions -
\emph{i.e.} a tradeoff between accuracy and a reduced number of
features.

An RDPG estimates the probability of observing interactions between
nodes (species) as a function of the nodes' latent variables, and is a
way to turn an SVD (which decompose one matrix into three) into two
matrices that can be multiplied to provide an approximation of the
network. The latent variables used for the RDPG, called the left and
right subspaces, are defined as
\(\mathscr{L} = \mathbf{U}\sqrt{\mathbf{\Sigma}}\), and
\(\mathscr{R} = \sqrt{\mathbf{\Sigma}}\mathbf{V}'\) -- using the full
rank of \(\mathbf{A}\), \(\mathscr{L}\mathscr{R} = \mathbf{A}\), and
using any smaller rank results in
\(\mathscr{L}\mathscr{R} \approx \mathbf{A}\). Using a rank of 1 for the
t-SVD provides a first-order approximation of the network. One advantage
of using an RDPG for the network reconstruction rather than an SVD is
that the number of components to estimate decreases; notably, one does
not have to estimate the singular values of the SVD. Furthermore, the
two subspaces can be directly multiplied to yield a network.

@@

It must be noted that despite bringing in a large amount of information
from the European species pool and interactions, the Canadian metaweb
has distinct structural properties. Following an approach similar to
Vermaat et al. (2009), we show in Supp. Mat. 3 that not only can we
observe differences in the multivariate space between the European and
Canadian metawebs, we can also observe differences in the same space
between random subgraphs from these networks. These results line up with
the studies spatializing metawebs that have been discussed in the
introduction: changes in the species pool are driving local structural
changes in the networks.

@@

\begin{figure}
\hypertarget{fig:inflation}{%
\centering
\includegraphics{figures/figure-validation.png}
\caption{Left: comparison of the probabilities of interactions assigned
by the model to all interactions (grey curve), the subset of
interactions found in GloBI (red), and in the Strong \& Leroux (2014)
Newfoundland dataset (blue). The model recovers more interactions with a
low probability compared to data mining, which can suggest that
collected datasets are biased towards more common or easy to identify
interactions. Right: distribution of the in-degree and out-degree of the
mammals from Canada in the reconstructed metaweb. This figure describes
a flat, relatively short food web, in which there are few predators but
a large number of preys.}\label{fig:inflation}
}
\end{figure}

@@

Second, we looked for all species in the Canadian pool known to the
Global Biotic Interactions (GloBI) database (Poelen et al., 2014), and
extracted their known interactions. Because GloBI aggregates observed
interactions, it is not a \emph{networks} data source, and therefore the
only information we can reliably extract from it is that a species pair
\emph{was reported to interact at least once}. This last statement
should yet be taken with caution, as some sources in GloBI (\emph{e.g.}
Thessen \& Parr, 2014) are produced through text analysis, and therefore
may not document direct evidence of the interaction. Nevertheless,
should the predictive model work, we would expect that a majority of
interactions known to GloBI would also be predicted. We retrieved 366
interactions between mammals from the Canadian species pool from GloBI,
33 of which were not predicted by the model; this results in a success
rate of 91\%. After performing this check, we set the probability of all
interactions known to GloBI to 1.

Finally, we downloaded the data from Strong \& Leroux (2014), who mined
various literature sources to identify trophic interactions in
Newfoundland. This dataset documented 25 interactions between mammals,
only two of which were not part of our (Canada-level) predictions,
resulting in a success rate of 92\%. These two interactions were added
to our predicted metaweb with a probability of 1. A comparison of
interaction densities for the inferred metaweb, and the Globi and
Newfoundland is shown in fig.~\ref{fig:inflation} and a table listing
all interactions in the predicted Canadian metaweb can be found in the
supplementary material.

@@

Because the confidence intervals on the inferred trait space are
probably over-estimates, we decided to apply a thresholding step to the
interactions after data inflation (see fig.~\ref{fig:thresholds} showing
the effect of varying the cutoff on \(P(i \rightarrow j)\)). Cirtwill \&
Hambäck (2021) proposed a number of strategies to threshold
probabilistic networks. Their methodology assumes the underlying data to
be tag-based sequencing, which represents interactions as co-occurrences
of predator and prey within the same tags; this is conceptually
identical to our Bernoulli-trial based reconstruction of a probabilistic
network. We performed a full analysis of the effect of various cutoffs,
and as they either resulted in removing too few interactions, or
removing enough interactions that species started to be disconnected
from the network, we set this threshold for a probability equivalent to
0 to the largest possible value that still allowed all species to have
at least one interaction with a non-zero probability. The need for this
slight deviation from the Cirtwill \& Hambäck (2021) methodology
highlights the need for additional development on network thresholding.

@@

Using a transfer learning framework we were able to construct a
probabilistic metaweb and (as per Dunne, 2006) is a list of potential
interactions, meaning that they will not necessarily be realized
wherever the two species co-occur. The t-SVD embedding is able to learn
relevant ecological features for the network. fig.~\ref{fig:degree}
shows that the first rank correlates linearly with generality and
vulnerability (Schoener, 1989), \emph{i.e.} the number of preys and
predators for each species. Importantly, this implies that a rank 1
approximation represents the configuration model for the metaweb,
\emph{i.e.} a set of random networks generated from a given degree
sequence (Park \& Newman, 2004). Accounting for the probabilistic nature
of the degrees, the rank 1 approximation also represents the \emph{soft}
configuration model (van der Hoorn et al., 2018). Both models are
maximum entropy graph models (Garlaschelli et al., 2018), with sharp
(all network realizations satisfy the specified degree sequence) and
soft (network realizations satisfy the degree sequence on average) local
constraints, respectively. The (soft) configuration model is an unbiased
random graph model widely used by ecologists in the context of null
hypothesis significance testing of network structure (\emph{e.g.}
Bascompte et al., 2003) and can provide informative priors for Bayesian
inference of network structure (\emph{e.g.} J.-G. Young et al., 2021).
It is noteworthy that for this metaweb, the relevant information was
extracted at the first rank. Because the first rank corresponds to the
leading singular value of the system, the results of
fig.~\ref{fig:degree} have a straightforward interpretation:
degree-based processes are the most important in structuring the
mammalian food web.

@@

\textbf{Acknowledgements:} We acknowledge that this study was conducted
on land within the traditional unceded territory of the Saint Lawrence
Iroquoian, Anishinabewaki, Mohawk, Huron-Wendat, and Omàmiwininiwak
nations. TP, TS, DC, and LP received funding from the Canadian Institute
for Ecology \& Evolution. FB is funded by the Institute for Data
Valorization (IVADO). TS, SB, and TP are funded by a donation from the
Courtois Foundation. CB was awarded a Mitacs Elevate Fellowship no.
IT12391, in partnership with fRI Research, and also acknowledges funding
from Alberta Innovates and the Forest Resources Improvement Association
of Alberta. M-JF acknowledges funding from NSERC Discovery Grant and
NSERC CRC. RR is funded by New Zealand's Biological Heritage Ngā Koiora
Tuku Iho National Science Challenge, administered by New Zealand
Ministry of Business, Innovation, and Employment. BM is funded by the
NSERC Alexander Graham Bell Canada Graduate Scholarship and the FRQNT
master's scholarship. LP acknowledges funding from NSERC Discovery Grant
(NSERC RGPIN-2019-05771). TP acknowledges financial support from NSERC
through the Discovery Grants and Discovery Accelerator Supplement
programs.
